# `behavr` tables{-} {#behavr}


## Variables and metavariables

When performing behavioural experiments, many animals are individually monitored.
For each of these individual, we record a set of **variables** (e.g. position, orientation, beam breaking,...).
Each variable is recorded at successive intervals, **over long durations**.
In addition to these variables, which change over time we also have **metavariables**.
Metavariables contain information **about each individual**.
This information can be defined (or recorded) before or after the experiment, but in all cases the point is that it **does not change over time** (e.g. treatment, genotype, lifespan).
Typically, metavariables are defined as extra *columns a the query file*, but they can also be computed.

For example, if we recorded `x` and `y` positions over time, for animals of different `sex` and `genotype`,
`x` and `y` would be the variables whilst `sex` and `genotype` would be the metavariables.


Regardless to what device was used to generate our data and how we loaded it, in `rethomics`, we will always have data and metadata linked together.


In order to handle efficiently large amounts of data (toghter with metadata), we have designed the `behavr` package.
A `behavr` table is essentially a [data.table](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) enhanced with metadata.
**When we load any behavioural data in `rethomics`, we get a `behavr` table as a result.**
  Therefore, it is important to understand how to manipulate and operate on this data structure.

## Toy data{-}

The `behavr` package has a set of functions to make toy data.
In order to have a look at a `behavr` object, lets create one.
First, we make a query:
  ```{r}
library(behavr)
query <- data.frame( experiment_id = "toy_experiment",
                     region_id = 1:10,
                     condition = c("A", "B") )
query
```

Then, we use `toy_dam_data()` to **simulate** one day of DAMS-like data for these ten animals (and two conditions):

  ```{r}
dt <- toy_dam_data(query, duration = days(1))
dt
```
As you can see, when we print `dt`, our `behavr` table, we have two fields: `METADATA` and `DATA`.
The former has only one row per animal and looks very much like query, with an extra column named `id`.
The later is the actual data of **all animals**. The special column `id` is also known as a key. It allows to map data to metadata.
In other words, there is a unique id for each individual.
In this specific example, `t` is the time and  `activity` is the number of beam crosses.

A quick way to retreive general information about a `behavr` table is to use `summary`:

  ```{r}
summary(dt)
```

This tells us immediately how many variables and metavariables we have.
It also tells us when are the first and last recording for each animal.


## Operating on `behavr` tables{-}

###  Variables{-}
Playing with variables is just like in `data.table`.
Read the [official data.table tutorial](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) for more functionalities.
For instance, we can add a new variable, `very_active` that is `TRUE` if and only if
there was at least two beam crosses in a minute:

  ```{r}
dt[, very_active := activity >= 2]
```
One can also filter data. For instance, we want to show only point before three hours:
  ```{r}
dt[t < hours(3)]
```

Note that `hours(3)` just converts 3h in seconds, because time is expressed in seconds in `rethomics` (there is also a `days()` and a `mins()` function).



###  Expand metavariables{-}

Sometimes, you would like to **use metavariables as if they were variables**.
Imagine, for instance, that we want to correct activity according to the condition, say we need to multiply activity by two for condition "A".
For this purpoe, we can use the function `xmv` (eXpand MetaVariable):


  ```{r}
dt[, activity_corrected := ifelse(xmv(condition) == "A", activity * 2, activity)]
```

If `condition` were a variable, we would use ` ifelse(condition == "A", activity * 2, activity)`.
Since it is in metadata, we just replace `condition` by `xmv(condition)`.


###  Metadata{-}

The Metadata is itself a table.
To perform operations on metadata, on can use `meta=TRUE` inside the `[]` operator.
For intace, only to display the metadata:

  ```{r}
dt[meta=TRUE]
```


We can use the same principle to add new metavariable:


  ```{r}
dt[,
   my_new_column := interaction(region_id, condition),
   meta=TRUE]
dt[meta=TRUE]
```

### Summarise data{-}

At some point, we will want to compute one or several summary variables for each animal and, for instance,
model these versus metavariable. For instance, we could compute the sum of all activity for each animal.
Since each animal is identified by its `id`, we use `by=id` for this:

  ```{r}
summary_dt <- dt[,
                 .(sum_activity = sum(activity)),
                 by=id]
summary_dt
```
Now, the data is only one point per animal, with `id`, and our new variable.
We aggreated the data, so it is quite managable to put the summary data and the metadata in the same table.
For that, we use `rejoin`:

  ```{r}
final_dt <- rejoin(summary_dt)
final_dt
```

This table is now perfectly suited for further statistical analysis and visualisation!


### Stitching{-}
TODO


